---
title: "R Programming — Control Structures, Functions & Apply Family (Class 2 — Continued)"
author: "Dr. Pacifique (notes prepared by James)"
date: "2025-10-05"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: default
  pdf_document: default
---

> Extracted and cleaned from the provided PDF. Fixed line-wrap artefacts and typos so all examples run in a fresh R session.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(1)
```

# Control structures

Control structures let you add *logic* to R programs so code can react to inputs or data. Common structures:

- `if` / `else`
- `for`
- `while`
- `repeat` (use `break` to exit)
- `break` (exit current loop)
- `next` (skip to next iteration)

## if / else

```{r}
x <- runif(1, 0, 10)

if (x > 3) {
  y <- 10
} else {
  y <- 0
}
y

# Equivalent as a single expression:
y2 <- if (x > 3) 10 else 0
y2
```

Vectorised `ifelse()` is handy for labeling results:

```{r}
pvalues <- c(.867, 0.0054, 0.0018, 0.1572, 0.0183, 0.5386)
results <- ifelse(pvalues < 0.05, "Significant", "Not significant")
results
```

## for loops

```{r}
for (i in 1:10) {
  print(i)
}

x <- c("a", "b", "c", "d")
for (i in 1:5) print(x[i])      # note: x[5] is NA

for (i in 1:5) print(1:i)
for (i in 5:1) print(1:i)
```

## while loops

```{r}
count <- 1
while (count < 10) {
  print(count)
  count <- count + 1
}

z <- 5
while (z >= 3 && z <= 10) {
  coin <- rbinom(1, 1, 0.5)
  if (coin == 1) {
    z <- z + 1
  } else {
    z <- z - 1
  }
}
z
```

## next and break

```{r}
for (i in 1:100) {
  print(1:i)
  if (i > 20) break
}
```

# Functions in R

Template:

```r
name <- function(parameters) {
  # statements
  return(value)
}
```

Examples (correcting math typos from the PDF):

```{r}
f1 <- function(x) {
  x^2 + 2
}
f1(4)

f2 <- function(x, y) {
  x^2 + y^2 - 4
}
f2(3, 5)

f <- function(x, y) {
  x + (2 * y) + 3
}
f(2, 3)

args(mean)        # see parameter names & defaults
```

# *apply family

Apply a function over rows/columns or elements of data structures.

```{r}
# Matrix example
m <- matrix(runif(12), nrow = 3)
m
log(m)
mean(m)                  # over all 12 elements

# Row/column summaries
rowMeans(m)
colSums(m)

# apply: margin = 1 (rows), 2 (columns)
apply(m, 1, mean)
apply(m, 2, mean)
```

## `lapply`, `sapply`, `vapply`

```{r}
# Fix a "sentinel" missing value and square numbers
fix_missing <- function(x) {
  x[x == -99] <- NA
  x
}
square <- function(x) x^2

df <- data.frame(a = c(1, -99, 3),
                 b = c(4, 5, -99),
                 c = c(-99, 2, 9))

# Apply to all columns and keep a data.frame
df[] <- lapply(df, fix_missing)
str(df)

# Transform a subset
df[1:2] <- lapply(df[1:2], square)
df
```

## `split`/`tapply` and multi-way `apply`

```{r}
# Titanic is a 4D contingency table
str(Titanic)

# Collapsing across dimensions
apply(Titanic, c(1, 2), sum)         # Class x Sex
apply(Titanic, c(1, 4), sum)[3:4, ]  # Class x Survived for 3rd & Crew
apply(Titanic, c(1, 2, 4), sum)[3:4, , ]

ft <- ftable(apply(Titanic, c(1, 4, 2), sum)[3:4, , ])
ft

# Column-wise proportions
prop.table(ft[1:2, ], margin = 2)
prop.table(ft[3:4, ], margin = 2)
```

# Mini project — student performance (standardise, score, grade)

```{r}
Student <- c("John Ncuti","Angela Bakame","Bruce Wizeye","Alexis Aganze",
             "Claude Rukundo","Joel Kagabo","Mary Ineza")
Math    <- c(600, 412, 358, 495, 512, 410, 522)
Science <- c(95,  99,  80,  82,  75,  89,  77)
English <- c(25,  22,  18,  20,  29,  30,  27)

roster <- data.frame(Student, Math, Science, English, stringsAsFactors = FALSE)

# Standardise subject scores (different scales)
z <- scale(roster[, c("Math","Science","English")])

# Performance score: row means of z-scores
score <- rowMeans(z)
roster <- cbind(roster, score)

# Grade by quantiles
cuts <- quantile(score, c(.8, .6, .4, .2))
roster$grade <- NA_character_
roster$grade[score >= cuts[1]] <- "A"
roster$grade[score <  cuts[1] & score >= cuts[2]] <- "B"
roster$grade[score <  cuts[2] & score >= cuts[3]] <- "C"
roster$grade[score <  cuts[3] & score >= cuts[4]] <- "D"
roster$grade[score <  cuts[4]] <- "F"

# Split full names into first/last (using last token as last name)
name_parts <- strsplit(roster$Student, " ")
lastname  <- vapply(name_parts, function(x) tail(x, 1), character(1))
firstname <- vapply(name_parts, function(x) paste(x[-length(x)], collapse = " "), character(1))

roster2 <- cbind(firstname, lastname, roster[, -1])
roster2 <- roster2[order(lastname, firstname), ]
roster2
```

# Aggregation & reshaping

## Transpose example

```{r}
cars <- mtcars[1:5, 1:4]
cars
t(cars)
```

## `aggregate()` by multiple keys

```{r}
options(digits = 3)
aggdata <- aggregate(mtcars,
                     by = list(cyl = mtcars$cyl, gear = mtcars$gear),
                     FUN = mean, na.rm = TRUE)
head(aggdata)
```

## Reshape with `reshape2` (melt/cast)

```{r, eval=FALSE}
# install.packages("reshape2")
library(reshape2)

id  <- c(1, 1, 2, 2)
time<- c(1, 2, 1, 2)
x1  <- c(5, 3, 6, 2)
x2  <- c(6, 5, 1, 4)
mydata <- data.frame(id, time, x1, x2)

m <- melt(mydata, id.vars = c("id","time"))
dcast(m, id ~ variable, mean)        # wide
dcast(m, id + time ~ variable, sum)  # tall
```

# Quick index / Table of contents (topics covered)
- Control structures (`if`/`else`, `for`, `while`, `break`, `next`)
- Functions and return values
- `apply`, `lapply`, `sapply`, `vapply`
- Titanic multi-way tables and proportions
- Student grading mini-project (standardisation + quantiles)
- Aggregation with `aggregate()`
- Reshaping with `reshape2`

```{r}
sessionInfo()
```
