---
title: "Data Visualization & Data Management — Class 2 (Extracted)"
author: "Dr. Pacifique (notes prepared by James)"
date: "2025-10-01"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: default
  pdf_document: default
---

> Extracted from the provided PDF and lightly cleaned so all code chunks run in a fresh R session.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Data Visualization and Data Management

## 1) Bar plots — toy COVID-19 monthly data

```{r}
# Small example dataset: confirmed by gender (two rows = Male/Female; columns = months)
data1 <- read.table(text='
Sept_2022 Oct_2022 Nov_2022 Dec_2022 Jan_2023 Feb_2023 Mar_2023 Apr_2023 May_2023 June_2023
1        4        6       30      140       67       47       23       39      134        80
2       10        2       29      164       72       81       28       63      211       132
', header = TRUE)

rownames(data1) <- c("Male","Female")
barplot(as.matrix(data1),
        main = "Confirmed cases vs Gender",
        beside = TRUE,
        col = c("gold3","red"))
legend("topright", c("Male","Female"), cex = 0.8, bty = "n",
       fill = c("gold3","red"))
```

```{r}
# Tests by gender (units arbitrary; two rows again = Male/Female)
data3 <- read.table(text='
Sept_2022 Oct_2022 Nov_2022 Dec_2022 Jan_2023 Feb_2023 Mar_2023 Apr_2023 May_2023 June_2023
1        867      582      1802     12003     3059      9205      5346      7269      5933       3245
2        842      436      1805     11785     3508     10908      4902      7883      5901       4646
', header = TRUE)

rownames(data3) <- c("Male","Female")
barplot(as.matrix(data3),
        main = "COVID-19 tests vs Gender",
        beside = TRUE,
        col = c("blue","purple"))
legend("topleft", c("Male","Female"), cex = 0.9, bty = "n",
       fill = c("blue","purple"))
```

```{r}
# Confirmed cases vs age categories
data2 <- read.table(text='
Sept_2022 Oct_2022 Nov_2022 Dec_2022 Jan_2023 Feb_2023 Mar_2023 Apr_2023 May_2023 June_2023
1        0        0         1         9        8         7         1         4        11         22
2       11        3        41       242      104        78        44        86       244        147
3        3        5        17        53       27        43         6        12        90         43
', header = TRUE)

rownames(data2) <- c("<15 yrs","15–49 yrs",">=50 yrs")
barplot(as.matrix(data2),
        main = "Confirmed cases vs Age categories",
        beside = TRUE,
        col = c("green","blue","purple"))
legend("topleft", rownames(data2), cex = 0.9, bty = "n",
       fill = c("green","blue","purple"))
```

## 2) Lines — confirmed vs treatment (two series)

```{r}
months <- c("Sept_2022","Oct_2022","Nov_2022","Dec_2022",
            "Jan_2023","Feb_2023","Mar_2023","Apr_2023","May_2023","June_2023")

confirmed <- c(14,8,59,304,139,128,51,102,345,212)
treatment <- c(12,0,26,217,104,57,12,47,23,20)

plot(confirmed, type = "l", pch = 21, col = "red", ylim = c(0, 350),
     xaxt = "n", ylab = "Number of patients",
     main = "Confirmed vs Treatment")
lines(treatment, col = "blue", lty = 2)
axis(1, at = 1:10, labels = months, las = 2, cex.axis = 0.8)
legend("topleft", legend = c("Confirmed","Received treatment"),
       cex = 0.8, bty = "n", lty = c(1,2), col = c("red","blue"))
```

## 3) Lines — number of tests

```{r}
number_of_tests <- c(1709,1018,3607,23788,6567,20113,10248,15152,11834,7891)
plot(number_of_tests, type = "o", col = "blue", ylim = c(0, 24000),
     xaxt = "n", ylab = "Number of Tests",
     main = "COVID-19 Tests")
axis(1, at = 1:10, labels = months, las = 2, cex.axis = 0.8)
```

# ggplot2 essentials

```{r}
# install.packages("ggplot2")  # if needed
library(ggplot2)

data(mtcars)

# Cylinder distribution (counts as factors)
ggplot(mtcars, aes(x = factor(cyl))) +
  geom_bar() +
  labs(title = "Distribution of car cylinder counts",
       x = "Number of cylinders", y = "Number of cars")

# Scatter with a linear model smoother
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Automobile Data", x = "Weight", y = "Miles per Gallon")

# Faceting example with grouping
mtcars2 <- mtcars
mtcars2$am <- factor(mtcars2$am, levels = c(0,1), labels = c("Automatic","Manual"))
mtcars2$vs <- factor(mtcars2$vs, levels = c(0,1), labels = c("V-Engine","Straight Engine"))
mtcars2$cyl <- factor(mtcars2$cyl)

ggplot(mtcars2, aes(x = hp, y = mpg, shape = cyl, color = cyl)) +
  geom_point(size = 2.5) +
  facet_grid(vs ~ am) +
  labs(title = "Automobile Data by Engine Type",
       x = "Horsepower", y = "Miles per Gallon")
```

# Basic data management (base R)

## cbind and rbind

```{r}
data2 <- data.frame(x = 1:3, y = c("a","b","c"))
str(data2)

cbind(data2, data.frame(z = 3:1))
rbind(data2, data.frame(x = 10, y = "z"))
```

## Create another variable (BMI) with a self‑contained dataset

```{r}
# Recreate the HEIGHT/WEIGHT data to avoid external file paths
data_class <- data.frame(
  HEIGHT = c(161,155,158,170,160,156,162,158,158,167,160,155,154,155,157,157,160,158,160,160,152,154,150,161,162,164,161,155,159,163,159,160,158,165,156,163,155,164,163,168,157,164),
  WEIGHT = c(50,49,42,65,60,52,58,46,45,51,60,42,53,48,48,48,53,52,51,53,44,56,63,52,57,49,52,54,46,50,61,55,45,63,60,56,52,47,52,55,48,58)
)

with(data_class, head(data.frame(HEIGHT, WEIGHT), 5))
tail(data_class, 3)

BMI <- with(data_class, WEIGHT / (HEIGHT/100)^2)
BMI <- round(BMI, digits = 1)

head(cbind(data_class, BMI))
tail(cbind(data_class, BMI), n = 10)

summary(BMI)
hist(data_class$HEIGHT, main = "Sample Histogram: HEIGHT")
hist(data_class$WEIGHT, main = "Sample Histogram: WEIGHT")
stem(BMI, scale = 2)
```

# Subsetting

```{r}
# Leadership toy dataset
manager <- c(1,2,3,4,5)
country <- c("US","US","UK","UK","UK")
gender  <- c("M","F","F","M","F")
age     <- c(32,45,25,39,99)  # 99 will be treated as missing
q1 <- c(5,3,3,3,2)
q2 <- c(4,5,5,3,2)
q3 <- c(5,2,5,4,1)
q4 <- c(5,5,5,NA,2)
q5 <- c(5,5,2,NA,1)

leadership <- data.frame(manager, country, gender, age, q1, q2, q3, q4, q5)
str(leadership)

# Fix the missing age code (99 -> NA) and create age categories
leadership$age[leadership$age == 99] <- NA
leadership$agecat <- NA_character_
leadership$agecat[!is.na(leadership$age) & leadership$age > 75] <- "Elder"
leadership$agecat[!is.na(leadership$age) & leadership$age >= 35 & leadership$age <= 75] <- "Middle Aged"
leadership$agecat[!is.na(leadership$age) & leadership$age < 35] <- "Young"

# One way to handle missing values: listwise deletion (na.omit)
is.na(leadership[, 5:9])
newdata <- na.omit(leadership)
newdata

# Sorting and typical subsets
order(leadership$age)               # order index
newdata2 <- leadership[order(leadership$age), ]
newdata3 <- leadership[, c("q1","q2","q3","q4","q5")]
newdata4 <- leadership[, -(1:2)]    # drop first two columns
newdata5 <- leadership[c(2,4,5), ]  # select specific rows
newdata6 <- subset(leadership, (age >= 35 | age < 24), select = c(q1,q2,q3,q4,q5))
newdata7 <- leadership[leadership$gender == "M" & leadership$age > 30, ]
```

# Data management with dplyr (nycflights13)

```{r}
# install.packages("nycflights13"); install.packages("dplyr")  # if needed
library(nycflights13)
library(dplyr)

str(flights)

# filter()
jan <- filter(flights, month == 1)
newyear <- filter(flights, month == 1, day == 1)
dec25 <- filter(flights, month == 12, day == 25)
jan_dec <- filter(flights, month %in% c(1, 12))
nov_dec <- filter(flights, month %in% c(11, 12))

# arrange()
data_10 <- arrange(flights, year, month, day)
top_delays <- arrange(flights, desc(arr_delay))
head(top_delays)

# select()
time_var <- select(flights, year, month, day)
ymd_only <- select(flights, year:day)
rest <- select(flights, -(year:day))

# mutate()
flights2 <- select(flights, year:day, ends_with("delay"), distance, air_time)
mutated <- mutate(flights2,
                  gain = arr_delay - dep_delay,
                  speed = distance / air_time * 60)
head(mutated)
```

