---
title: "R for Data Science — Class 1 (Extracted & Knit-Ready)"
author: "Dr. Pacifique (notes prepared by James)"
date: "2025-09-14"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: default
  pdf_document: default
---

> Extracted from class notes. Code adjusted where necessary to be runnable in an R Markdown environment.

# Important definitions in data science

- **Statistics**
- **Modelling**
- **Computation**
- **Calculus**

# Today’s class — Agenda

- Installation of R *(reference)*
- Installation of RStudio *(reference)*
- Calculator in R
- Data importation in R
- Data description using numerical measures and graphs
- Data structures (Vectors, Factors, Lists, Data frame, Matrix and Arrays)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Vectors — The foundational R data structure

## Numeric
```{r}
x_num <- c(0.5, 0.6)
age <- c(20, 35, 32, 29)
summary(age)

age <- c(35, 24, 18, 24)
mean(age)
```

## Logical
```{r}
x_log1 <- c(TRUE, FALSE)
x_log2 <- c(T, F)
```

## Character
```{r}
x_char <- c("a", "b", "c")
class_vec <- c("M","F","F","M")
```

## Integer
```{r}
x_int <- 9:29
```

## Complex
```{r}
x_cx <- c(1+0i, 2+4i)
```

# Matrices and basic operations

```{r}
x <- matrix(1:6, nrow = 2, ncol = 3)
x

(y <- matrix(1:6, nrow = 3, ncol = 2))

x %*% y
```

# Lists

```{r}
x_list <- list(1, "a", TRUE, 1 + 4i)
x_list
```

# Factors

```{r}
x_fac <- factor(c("yes", "yes", "no", "yes", "no"))
x_fac
```

# Missing values (NA)

```{r}
x <- c(1, 2, NA, 10, 3)
is.na(x)

x2 <- c(1, 2, 4, "NA", 5)
bad <- is.na(x2)
bad
x2[!bad]
```

What if there are multiple objects and you want only the rows with no missing values in any of them?

```{r}
x <- c(1, 2, NA, 4, NA, 5)
y <- c("a", "b", NA, "d", NA, "f")
good <- complete.cases(x, y)
good
```

# Coercion

- If a **character** is present in a vector, R converts everything to **character**.
- If a vector only contains **logical** and **numeric**, R converts logicals to numbers (`TRUE` → 1, `FALSE` → 0).

```{r}
sum(c(TRUE, TRUE, FALSE, FALSE, FALSE))
```

# Creating data frames

Data frames can contain mixed types (numeric, character, etc.).

```{r}
studentID <- c(1, 2, 3, 4, 5)
math_score <- c(12, 17, 10, 9, NA)
gender <- c("M","F","M","M","F")
it_score <- c(13, 18, 11, 10, 19)
scoredata <- data.frame(studentID, gender, math_score, it_score)
scoredata
# View(scoredata)  # interactive; disabled in Rmd
```

## Creating data via editor (interactive; for reference)

```{r, eval=FALSE}
data_class2 <- data.frame(height=numeric(0), weight=numeric(0), bmi=numeric(0))
data_class2 <- edit(data_class2)
```

# Data importation (illustrations)

R can import from Excel, statistical packages (SAS/SPSS/Stata), text (CSV/TSV/JSON/XML), and databases (SQL/MySQL/Oracle/Access).

```{r, eval=FALSE}
# Example: text file (adjust path)
data_class <- read.table("C:/path/to/class_data.txt", header = TRUE)
variable.names(data_class)
head(data_class)
tail(data_class)
data_class[10:20, ]
summary(data_class$WEIGHT)
length(data_class$WEIGHT)
data_class[, -1]
```

# Built-in datasets

```{r}
data("airquality")
variable.names(airquality)
str(airquality)
```

# Exploratory data analysis (EDA)

For the following, we’ll reconstruct `data_class` to make the notebook self-contained.

```{r}
data_class <- data.frame(
  HEIGHT = c(161,155,158,170,160,156,162,158,158,167,160,155,154,155,157,157,160,158,160,160,152,154,150,161,162,164,161,155,159,163,159,160,158,165,156,163,155,164,163,168,157,164),
  WEIGHT = c(50,49,42,65,60,52,58,46,45,51,60,42,53,48,48,48,53,52,51,53,44,56,63,52,57,49,52,54,46,50,61,55,45,63,60,56,52,47,52,55,48,58)
)
data_class
```

## Summary statistics

```{r}
summary(data_class)
apply(data_class, 2, mean)
apply(data_class, 2, sd)

c(Mean = mean(data_class$HEIGHT), SD = sd(data_class$HEIGHT))
c(Mean = mean(data_class$WEIGHT), SD = sd(data_class$WEIGHT))
```

## Variation & covariation (numeric and graphs)

### Categorical vs continuous — `iris`

```{r}
data("iris")
head(iris)

mean(iris$Petal.Length)
by(iris$Petal.Length, iris$Species, mean)
by(iris$Petal.Length, iris$Species, sd)
by(iris$Petal.Length, iris$Species, summary)

boxplot(Sepal.Length ~ Species, data = iris, main = "Comparison")
```

### Two categorical variables — `UCBAdmissions`

```{r}
ucba <- data.frame(UCBAdmissions)
head(ucba)

cross <- xtabs(Freq ~ Gender + Admit, data = ucba)
cross

# Proportions by column (admission outcome)
prop.table(cross, 2)

# Simpson's paradox illustration — Department A only
cross_A <- xtabs(Freq ~ Gender + Admit, data = subset(ucba, Dept == "A"))
prop.table(cross_A, 1)
```

# Bar charts

```{r}
dat1 <- read.table(text ="ProdA ProdB ProdC ProdD
1 110 50 60 70
2 120 50 80 65", header= TRUE)

barplot(as.matrix(dat1), beside = FALSE)

dat2 <- read.table(text = "A B C D E F G
1 10 80 30 90 70 60 90
2 20 50 70 50 40 10 40
3 60 80 80 60 60 30 160
4 20 40 70 80 20 10 70", header = TRUE)

barplot(as.matrix(dat2))
```

# Scatter plot & correlation

```{r}
plot(WEIGHT ~ HEIGHT, data = data_class)
cor(data_class$HEIGHT, data_class$WEIGHT)
```

# Linear models

```{r}
plot(WEIGHT ~ HEIGHT, data = data_class)
abline(lm(WEIGHT ~ HEIGHT, data = data_class))

data_lm <- lm(WEIGHT ~ HEIGHT, data = data_class)
summary(data_lm)
```

# Histograms

```{r}
hist(data_class$HEIGHT)
hist(data_class$WEIGHT)
```

# Linear fit vs smoother

```{r}
plot(WEIGHT ~ HEIGHT, data = data_class)
abline(lm(WEIGHT ~ HEIGHT, data = data_class))
lines(lowess(data_class$HEIGHT, data_class$WEIGHT))
```

# Data management — Create another variable (BMI)

```{r}
BMI <- with(data_class, WEIGHT / (HEIGHT/100)^2)
BMI <- round(BMI, digits = 2)

head(cbind(data_class, BMI), 5)
tail(cbind(data_class, BMI), 10)
```

## Summary of BMI

```{r}
summary(BMI)
stem(BMI, scale = 2)
```
