---
title: "Regression, Simulation & Visualization — Class 4"
author: "Dr. Pacifique (notes prepared by James)"
date: "2025-10-08"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: default
  pdf_document: default
---

> Extracted and cleaned from class materials. All code is runnable in RStudio.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(123)
```

# Regression

Regression analyses the relationship between a **dependent** (outcome) variable and one or more **independent** (predictor) variables.

## Types of Regression

| Type | Description |
|------|--------------|
| Simple Linear | Predicts a quantitative response from one quantitative explanatory variable |
| Polynomial | Uses powers of a quantitative variable to model nonlinear trends |
| Multiple Linear | Predicts a quantitative response from two or more explanatory variables |
| Multilevel | For hierarchical data (e.g., students within schools) |
| Multivariate | Predicts multiple responses from one or more explanatory variables |
| Logistic | Predicts a categorical (binary) outcome |
| Poisson | For count data |
| Cox Proportional Hazards | For survival or time-to-event data |
| Time-Series | For correlated temporal data |
| Nonlinear | For inherently nonlinear relationships |
| Nonparametric | For relationships determined from data shape, not predefined equations |
| Robust | Resistant to outliers |

## Linear Model Basics

- Used to predict a continuous variable  
- Expressed as: \( y = b_0 + b_1x + e \)
- Residuals = prediction errors  
- RSS = Residual Sum of Squares  
- MSE = Mean Square Error  
- RMSE = Root Mean Square Error  

“All models are wrong, but some are useful.” — George Box

# OLS Regression in R

```{r}
data("women")
fit <- lm(weight ~ height, data = women)
summary(fit)
```

```{r}
plot(women$height, women$weight,
     xlab = "Height (in inches)",
     ylab = "Weight (in pounds)")
abline(fit, col = "blue", lwd = 2)
```

# Polynomial Regression

```{r}
fit2 <- lm(weight ~ height + I(height^2), data = women)
summary(fit2)

plot(women$height, women$weight,
     xlab = "Height (in inches)",
     ylab = "Weight (in pounds)")
lines(women$height, fitted(fit2), col = "red", lwd = 2)
```

# Regression Diagnostics

```{r}
par(mfrow = c(2, 2))
plot(fit)
```

Assumptions checked:

1. **Normality** – residuals follow a normal distribution  
2. **Independence** – errors are independent  
3. **Linearity** – relationship between predictors and response is linear  
4. **Homoscedasticity** – variance of residuals is constant  

If violated, consider transformations, adding terms, or other model types.

# Multiple Linear Regression (Exercise)

```r
# Homework: Find a dataset, fit multiple linear regression, and interpret results
```

# Example — mtcars

```{r}
model <- lm(mpg ~ wt, data = mtcars)
summary(model)

plot(mpg ~ wt, data = mtcars)
abline(model, col = "darkgreen", lwd = 2)
```

Interpretation: For every 1-unit increase in weight, miles per gallon decreases by about 5.3.

```{r}
predict(model, newdata = data.frame(wt = 6))
```

# Simulation in R

Simulation introduces randomness — useful for bootstrapping, testing, and modeling.

## Random Numbers

```{r}
rnorm(5)
set.seed(1)
rnorm(5)
```

## Simulating a Linear Model

```{r}
set.seed(20)
x <- rnorm(100)
e <- rnorm(100, 0, 2)
y <- 0.5 - 2 * x + e

plot(x, y, main = "Simulated Linear Model")
```

# Sampling

```{r}
set.seed(1)
sample(1:10, 4)
sample(letters, 5)
sample(1:10, replace = TRUE)
```

# Data Visualization

```{r}
country <- c("Australia", "Austria", "Belgium","Canada","Denmark","Finland","France","Germany","Greece","Ireland","Italy","Japan","Netherland","New Zealand","Norway","Portugal","Spain","Sweden","Switzerland","UK","USA")
Income.inequality <- c(7.0,4.8,4.6,5.6,4.3,3.7,5.6,5.2,6.2,6.0,6.7,3.4,5.3,6.8,3.9,8.0,5.5,4.0,5.7,7.2,8.6)
Index.HS <- c(0.07,0.01,-0.23,-0.07,-0.19,-0.43,0.05,-0.06,0.38,0.25,-0.12,-1.26,-0.51,0.29,-0.63,1.18,-0.30,-0.83,-0.46,0.79,2.02)
GDP_WB <- c(45926,47682,43435,45066,45537,30676,39328,46401,26851,49393,35463,36319,48253,37679,65615,28760,33629,45297,59540,40233,54630)

data.21 <- data.frame(country, Income.inequality, Index.HS, GDP_WB)

plot(data.21[c("Income.inequality","Index.HS")], pch = 20, col = "red")
text(data.21$Income.inequality, data.21$Index.HS, labels = data.21$country, pos = 4, cex = 0.7)
title(main = "Income Inequality vs Health & Social Index",
      xlab = "Income Inequality (5th Ratio)",
      ylab = "Index of Health & Social Problems")
text(x = 5, y = 1.5, labels = paste("r =", round(cor(data.21$Income.inequality, data.21$Index.HS), 2)))
```

# Linear Fit Overlay

```{r}
lm.ineq <- lm(Index.HS ~ Income.inequality, data = data.21)
abline(lm.ineq, col = "blue", lwd = 2)
```

# Advanced Example — Random Pathway Simulation

```{r}
set.seed(1002)
n <- 30
p <- 5
z <- matrix(runif(n * p, 0, 1), nrow = n, ncol = p)
x <- 3 * cos(z[,1]) + 2 * rnorm(n)
x <- as.matrix(x)
beta.true <- rep(1, ncol(x))

hfun1 <- function(zvec) (10 * cos(zvec[1]) - 15 * (zvec[2])^2 + 10 * exp(-zvec[3]) * zvec[4] - 8 * sin(zvec[5]) * cos(zvec[3]) + 20 * (zvec[1] * zvec[5]))
h1 <- apply(z, 1, hfun1)
eps <- rnorm(n)
y <- x * beta.true + h1 + eps

plot(y ~ x, main = "Simulated Pathway Relationship")
```

```{r}
sessionInfo()
```
